Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-05-03 Rmarkdown/wgbs_downstream_analyses.Rmd
Edit to WGBS downstream analysis

A barrier to analyses based on m-tuples has been a lack of software. To overcome this barrier, I develop \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples. \texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples. I describe methods available in \texttt{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
\end{chapabstract}

@@

All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$

@@

Unfortunately, early experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or pooled the replicates prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}). Simply pooling replicates and analysing as if they were a single sample ignores all within-group variability and should not be used.

Technical variability is ideally orthogonal to the biological variability, but this rarely occurs in practice. Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}. Again, this emphasises the importance of good experimental design such as randomisation, replication and the use of positive and negative controls.

@@

While there have been reports of DMCs resulting in a phenotypic difference \citep[e.g.,]{Furst:2012gh}, DMCs are mostly tested as a prelude to the identification of differentially methylated regions (DMRs). Moreover, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome bisulfite-sequencing experiments. This will remain true while sample sizes can be counted on one or two hands and the average sequencing depth is $10\times$ to $30\times$.

@@

`BiSeq` \citep{Hebestreit:2013ko} uses a hierarchical procedure to test for differential methylatoin at _a priori_ defined regions. To begin, `BiSeq` first defines _CpG clusters_ by identifying CpGs that are within a user-specified genomic distance of one another and that have sufficient sequencing coverage across the set of samples[^biseq_alternative]. While there is clearly a 'data-driven' component to these cluster definitions, I reserve the use of 'data-driven regions' for those which are based on the methylation levels of loci rather than their genomic co-ordinates.

@@

Methods for identifying _data-driven_ DMRs are fairly _ad hoc_. The most common approach is to scan the genome for clusters of DMCs and declare these to be DMRs. The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e. below genome-wide significance). Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.

@@

\citet{Jaffe:2011hy} first developed formal statistical tests for differential variability of methylation for both the one-group and two-group experiments. These methods were developed for us with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}). In a one-group experiment, a variably methylated region is one that has increased variability compared to 'similar' regions elsewhere in the genome. In a two-group experiment, a differentially variable locus is one where the variation in one group is significantly larger than that in the other group.

Statistical tests of variances are well known to be more difficult than tests of means and require larger sample sizes. A more subtle difficulty is in dealing with outliers. An outlier in one group will greatly increase the variation in that group, but this does not necessarily mean that the locus is differentially variably methylated. It might, for example, be due to an error in the assay. It is not difficult to envisage an example where the two groups in fact have very similar variability when the outlier is excluded.

@@

All these methods are based on analysing $\mathbf{m}$ and $\mathbf{u}$ or $\bm{\beta}$.

@@

Methylation data in the form of the vector of $\beta$-values, $\bm{\beta}$, may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}. One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs). Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.

@@

In this section I review the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples. This extra information is only available from sequencing-based assays. I also describe existing software that implement some of these methods.

@@

Broadly speaking, on the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero. On the other hand, if we observe all possible $2^{m}$ methylation patterns at equal frequency, then the m-tuple has the maximum methylation entropy (typically normalised to one). Depending on the frequency of the observed methylation patterns patterns, we obtain intermediate values of the methylation entropy.

@@

While straightforward, this approach is also obviously limited to a subset of the methylation loci in the genome. \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest. These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data. Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.

@@

Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom. Even then, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}. In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}. The poster child for the potential importance of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}. Genetically identical mice with different versions of the $A^{vy}$ allele are phenotypically distinct. Those mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic, and have an increased susceptibility to tumours; those mice with a methylated version of the allele have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.

@@

[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size. While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results.

@@

`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias. In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.

`MPFE` is designed to estimate the frequency of epialleles by maximising a multinomial likelihood that includes error terms for both incomplete bisulfite-conversion and sequencing error. The maximisation of this likelihood is computatationally demanding, as evidenced by the need for a "fast" algorithm that approximates the likelihood. `MPFE` is designed for amplicon bisulfite-sequencing and may not scale to whole-genome data. The input is a file containing the number of times each methylation pattern was observed at that m-tuple. Unfortunately, `MPFE` does not provide a way to create this table.

@@

I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations. Two give two examples, `DMEAS` \citep{He:2013cj} is only available as a Windows binary or as a Perl script that itself is only available as a PDF file and I have been unable to install `methclone` due to compilation errors.

@@

[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [https://github.com/PeteHaitch/MethylationTuples](https://github.com/PeteHaitch/MethylationTuples).

Bioconductor has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. This makes it easy to use for users already familiar with the `GenomicRanges` package. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e. \texttt{type = `equal'}). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing methylation patterns at 3-tuples for $n$ samples. The similarities to the output format of `methtuple` are clear[^methtuple_output], with the added advantage that a single `MethPat` object can contain data from multiple samples.

@@

The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the `[` method. It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods. Several other useful utility functions for working with `MethPat` objects include:

@@

Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the _EPISCOPE_ whole-genome bisulfite-sequencing data with various sized m-tuples. Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample). However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.

\begin{table}[h]
\centering
\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data. All m-tuples are stranded. The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The `number of rows' corresponds to the number of m-tuples in the object. The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples. The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\end{table}

>>>>

A barrier to analyses based on m-tuples has been a lack of software. To help eliminate this barrier, I develop \texttt{MethylationTuples}, an R package for managing, analysing and visualising methylation patterns at m-tuples. \texttt{MethylationTuples} complements \texttt{methtuple} (Chapter \ref{chap:wgbs_bioinformatics_analysis}) by providing a framework for the manipulation and analysis of methylation patterns at m-tuples. I describe methods available in \texttt{MethylationTuples} for the downstream analysis of whole-genome bisulfite-sequencing data.
\end{chapabstract}

@@

@@

Initial experiments with whole-genome bisulfite-sequencing rarely had replicates of any kind or these were pooled prior to analysis (e.g., the analysis of the H1 and IMR90 cell lines in \citet{Lister:2009hy}). Simply pooling replicates and analysing as if they were a single sample ignores all within-group variability and should not be used.

Technical variability is ideally orthogonal to the biological variability, but this rarely occurs in practice. Indeed, high-throughput sequencing experiments are particularly susceptible to batch effects, and other sources of unwanted variation, that can swamp the biological variation of interest \citep{Leek:2010jq}. This again emphasises the importance of good experimental design, with randomisation, replication and the use of positive and negative controls.

@@

While there have been reports of DMCs resulting in a phenotypic difference \citep[e.g.,][]{Furst:2012gh}, DMCs are mostly tested as a prelude to the identification of differentially methylated regions (DMRs). Moreover, with approximately 25 million CpGs in the human genome, not to mention the many, many more non-CpG cytosines, it is an optimist who aims for the reliable detection of DMCs from whole-genome bisulfite-sequencing experiments. This will remain true while sample sizes can be counted on one or two hands and the average sequencing depth is $10\times$ to $30\times$.

@@

`BiSeq` \citep{Hebestreit:2013ko} uses a hierarchical procedure to test for differential methylatoin at _a priori_ defined regions. To begin, `BiSeq` first defines _CpG clusters_ by identifying CpGs that are within a user-specified genomic distance of one another and that have sufficient sequencing coverage across the set of samples[^biseq_alternative]. While there is clearly a 'data-driven' component to these cluster definitions, I reserve the use of 'data-driven regions' for those that are based on the methylation levels of loci rather than their genomic co-ordinates.

@@

Methods for identifying _data-driven_ DMRs are statistically _ad hoc_. The most common approach is to scan the genome for clusters of DMCs and declare these to be DMRs. The initial scan for DMCs will typically use a relaxed statistical significance threshold (i.e. not necessarily genome-wide significant). Notably, many of these methods do not include a formal statistical test of differential methylation at the region-level \citep[e.g.,][]{Lister:2009hy, Lister:2011kg, Hansen:2011gu, Feng:2014iq}.

@@

\citet{Jaffe:2011hy} first developed formal statistical tests for differential variability of methylation for both the one-group and two-group experiments. These methods were developed for use with data from the CHARM array (see \ref{sec:enzyme_digestion_assays}). In a one-group experiment, a variably methylated region is one that has increased variability compared to 'similar' regions elsewhere in the genome. In a two-group experiment, a differentially variable locus is one where the variation in one group is significantly larger than that in the other group.

Statistical tests of variances are well known to be more difficult than tests of means and require larger sample sizes. A more subtle difficulty is in dealing with outliers. An outlier in one group will greatly increase the variation in that group, but this does not necessarily mean that the locus is differentially variably methylated. It might, for example, be due to an error in the assay. It is not difficult to envisage an example where the two groups in fact have very similar variability once the outlier is excluded.

@@

@@

Methylation data in the form of the vectors of $\mathbf{m}$ and $\mathbf{u}$  may also be used by methods to segment the genome or epigenome into regulatory regions \citep{Stadler:2011kt}. One type of region that has received particular attention are so-called _partially methylated domains_ (PMDs). Partially methylated domains are long stretches of the genome where the average methylation level, $B_{i}$, is away from the extremes of $0$ and $1$, typically in the range $0.2$ to $0.7$.

@@

This section reviews the different types of questions that can be addressed by expanding our analysis to use methylation patterns at m-tuples (m $> 2$) rather than just 1-tuples. This extra information is only available from sequencing-based assays. We also review existing software that implement some of these methods.

@@

On the one hand, if we only observe a single unique methylation pattern, then the m-tuple has the minimum methylation entropy of zero. On the other hand, if we observe all possible $2^{m}$ methylation patterns at equal frequency, then the m-tuple has the maximum methylation entropy (typically normalised to one). Depending on the frequency of the observed methylation patterns patterns, we obtain intermediate values of the methylation entropy.

@@

While straightforward, this approach is also obviously limited to the small number of methylation loci that are nearby to a heterozygous genetic variant. \citet{Fang:2012ci} and \citet{Peng:2012dh} published methods to detect allele-specfic methylation that do not require heterozygous genetic variants nearby to the methylation locus of interest. These methods rely on the probabilistic assignment of reads to alleles, which are treated as missing data. Unfortunately, there is no publicly available software implementing the method proposed by \citet{Peng:2012dh} and so I do not discuss it further.

@@

Most of the examples of epialleles with a phenotypic consequence come from the plant kingdom and, even there, the number of such epialleles is small: a review from 2012 put the number at "about a dozen" \citep{Weigel:2012fj}. In mammals, the study of epialleles has focused on identifying _metastable_ epialleles, which are epialleles that are mitotically heritable \citep{Rakyan:2002wz}. The poster child for the potential importance of epialleles in mammals is the Agouti viable yellow ($A^{vy}$) allele \citep{Morgan:1999ds}. Genetically identical mice with different versions of the $A^{vy}$ allele are phenotypically distinct. Those mice with an unmethylated version of the allele have a yellow coat, are obese, diabetic, and have an increased susceptibility to tumours; those mice with a methylated version of the allele have a _pseudoagouti_[^pseudoagouti_vs_agouti] (brown) coat and none of the associated health defects.

@@

[^sampling_with_replacement]: Strictly speaking, this is in fact an estimate of the rate of epipolymorphism of the locus under a model that assumes sampling with replacement or, equivalently, an infinite population size. While neither assumption is true, the correction for sampling without replacement from a finite population will not substantially affect the results provided that the sequencing depth is high.

@@

`methclone` uses the observed frequencies of methylation patterns as being unbiased estimates of the true epiallele frequencies and does not attempt to account for potential sources of bias. In contrast, `MPFE` ([http://www.bioconductor.org/packages/release/bioc/html/MPFE.html](http://www.bioconductor.org/packages/release/bioc/html/MPFE.html), [http://f1000.com/posters/browse/summary/1097258](http://f1000.com/posters/browse/summary/1097258)), an R/Bioconductor package for "[estimating] the distribution of methylation patterns [i.e. epialleles]" at m-tuples, uses a probabilistic model to account for some of these biases.

`MPFE` is designed to estimate the frequency of epialleles by maximising a multinomial likelihood that includes error terms for both incomplete bisulfite-conversion and sequencing error. The maximisation of this likelihood is computatationally demanding, as evidenced by the need for a "fast" algorithm that approximates the likelihood. `MPFE` is designed for amplicon bisulfite-sequencing and may not scale to whole-genome data. The input is a file containing the number of times each methylation pattern was observed at that m-tuple. Unfortunately, `MPFE` does not provide a way to create this file.

@@

I have also experienced considerable difficulty in applying some of these methods owing to poor software implementations. To give two examples, `DMEAS` \citep{He:2013cj} is only available as a Windows binary or as a Perl script that itself is only available as a PDF file, and I have been unable to install `methclone` due to compilation errors.

@@

[^bioc]: `MethylationTuples` has not yet been submitted to Bioconductor but its development is being published to [bioconductor.org/packages/GenomicTupless](bioconductor.org/packages/GenomicTuples).

Bioconductor already has excellent support for working with data defined on genomic ranges via the `IRanges` and `GenomicRanges` packages \citep{Lawrence:2013hi, Lawrence:2014gy}. However, genomic tuples, such as the co-ordinates of an m-tuples, do not naturally fit into this framework[^tuples_vs_ranges]. Therefore, I first wrote a Bioconductor package for working with genomic tuples, rather unimaginatively called `GenomicTuples`, first released as part of Bioconductor version 3.0 ([http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html](http://www.bioconductor.org/packages/3.0/bioc/html/GenomicTuples.html)). In fact, `GenomicTuples` is heavily based on the `GenomicRanges` package, with modifications for tuple-specific operations. This makes it easy to use for users already familiar with the `GenomicRanges` package. For example, there is a tuple-specific method for the `findOverlaps` generic function to identify genomic tuples with equal co-ordinates (i.e. \texttt{type = `equal'}). Since the classes in `GenomicTuples` extend those defined in `GenomicRanges`, these have excellent interoperability with existing Bioconductor infrastructure.

@@

Figure \ref{fig:MethPat} is a schematic of a `MethPat` object storing methylation patterns at 3-tuples for $n$ samples. The similarities to the output format of `methtuple` are clear (see Figure \ref{fig:3-tuples}), with the added advantage that a single `MethPat` object can contain data from multiple samples.

@@

The `MethPat` object provides fast subsetting by rows (m-tuples) and columns (samples) via the "`[`" method. It also benefits from fast subsetting based on overlaps of m-tuples with genomic features via the `findOverlaps()`-based methods. Several other useful utility functions for working with `MethPat` objects include:

@@

Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the _EPISCOPE_ whole-genome bisulfite-sequencing data with various sized m-tuples. We see that as the size of the m-tuples increases, the data become sparser: most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample). However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.

\begin{table}[h]
\centering
\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data. All m-tuples are stranded. The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The `number of rows' corresponds to the number of m-tuples in the object. The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples. The final column is a measure of how sparse the data are: a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\end{table}
